Value Ordering, Discrepancies, and Restarts in Subgraph Solvers: \\ Some Half-Baked Mucking Around

### 2017-10-04 paper/paper.tex
A bit cleaner
More things

\newcommand{\siplabel}[1]{\label{line:sip:#1}}
\newcommand{\siplineref}[1]{line~\ref{line:sip:#1}}
\begin{algorithm}[p]\DontPrintSemicolon\small\SetInd{0.1em}{1em}
    \begin{tikzpicture}[remember picture,overlay]
        \coordinate (restart1sc) at ($(pic cs:restart1s) + (0, 0.15)$);
        \coordinate (restart1ec) at ($(pic cs:restart1e) + (0, 0.00)$);
        \node [fill=magma8, rounded corners, fit=(restart1sc) (restart1ec)] { };
        \coordinate (restart2sc) at ($(pic cs:restart2s) + (0, 0.15)$);
        \coordinate (restart2ec) at ($(pic cs:restart2e) + (0, 0.00)$);
        \node [fill=magma8, rounded corners, fit=(restart2sc) (restart2ec)] { };
        \coordinate (restart3sc) at ($(pic cs:restart3s) + (0, 0.15)$);
        \coordinate (restart3ec) at ($(pic cs:restart3e) + (0, 0.00)$);
        \node [fill=magma8, rounded corners, fit=(restart3sc) (restart3ec)] { };
        \coordinate (restart4sc) at ($(pic cs:restart4s) + (0, 0.15)$);
        \coordinate (restart4ec) at ($(pic cs:restart4e) + (0, 0.00)$);
        \node [fill=magma8, rounded corners, fit=(restart4sc) (restart4ec)] { };
        \coordinate (restart5sc) at ($(pic cs:restart5s) + (0, 0.15)$);
        \coordinate (restart5ec) at ($(pic cs:restart5e) - (0, 0.10)$);
        \coordinate (restart5rc) at ($(pic cs:restart5r) + (0, 0.00)$);
        \node [fill=magma8, rounded corners, fit=(restart5sc) (restart5ec) (restart5rc)] { };
        \coordinate (restart6sc) at ($(pic cs:restart6s) + (0, 0.15)$);
        \coordinate (restart6ec) at ($(pic cs:restart6e) + (0, 0.00)$);
        \node [fill=magma8, rounded corners, fit=(restart6sc) (restart6ec)] { };
        \coordinate (restart7sc) at ($(pic cs:restart7s) + (0, 0.15)$);
        \coordinate (restart7ec) at ($(pic cs:restart7e) + (0, 0.00)$);
        \node [fill=magma8, rounded corners, fit=(restart7sc) (restart7ec)] { };
        \coordinate (restart8sc) at ($(pic cs:restart8s) + (0.05, 0.15)$);
        \coordinate (restart8ec) at ($(pic cs:restart8e) + (0, 0.00)$);
        \node [fill=magma8, rounded corners, fit=(restart8sc) (restart8ec)] { };
    \end{tikzpicture}

@@

        \nl \lElseIf{$S = \KwSty{false}$}{$\KwSty{return}~\KwSty{false}$}

@@

\medskip

@@

For value ordering, which occurs on \siplineref{valueordering} of \cref{algorithm:sip}, we will
compare four strategies:

@@

\paragraph{Tailored} We select vertices from highest degree to lowest degree. This is done by
permuting the target graph at the top of search. This heuristic is the current state of the art.

\paragraph{Anti} We select vertices from lowest degree to highest degree. This heuristic is used as
a sanity check: we expect tailored to beat random, and random to beat anti.

\paragraph{Biased} As with the random heuristic, we shuffle the vertices before we enter the
$\KwSty{foreach}$ loop. Instead of picking a vertex uniformly at random, we use a weight of
$2^{\left|\neighbourhood(v)\right|}$ when deciding whether to branch on vertex $v$ next. (We cheat
slightly to avoid range issues with random number generation: we give the vertex with highest degree
in the entire graph weight $2^{50}$, and then give lower degree vertices smaller powers, giving
every vertex weight at least 1.  This is a bit of hack but programming is hard.) This heuristic
heavily prefers vertices of high degree, similar to the tailored heuristic, but will give the same
chance of being selected next to two vertices of equal degree. This heuristic also introduces an
element of randomness, and so can be used with restarts.

@@

In \cref{figure:value-ordering-heuristics} we show the cumulative number of instances solved over
time using the four different value-ordering heuristics. As expected, the tailored heuristic is much
better (?? give aggregate speedups for both all and just satisfiable instances) than random
ordering, which is in turn ?? times better than the anti-heuristic.

@@

    \caption{An instance by instance comparison of the tailored and biased heuristics.}
    \label{figure:scatter-heuristics}
\end{figure}

The performance of the biased and tailored heuristics is very similar: the biased heuristic has
slightly worse performance below a 100s timeout, and effectively equal performance above it. This
shows that we can introduce an element of randomness into the tailored value-ordering heuristic
without adversely affecting its performance \emph{in aggregate}. \Cref{figure:scatter-heuristics}
gives a more detailed look at these results. In this plot, satisfiable and unsatisfiable instances
are shown using dots and crosses respectively. On the $x$-axis we measure the tailored heuristic,
whilst on the $y$-axis we measure random + restarts. Points on the outer axes represent instances
which timed out with one heuristic but not the other. This plot shows that on a case by case basis,
randomness can make a large difference to individual satisfiable instances.

@@

In \cref{figure:restarts} we show the effects of adding restarts to the algorithm. With restarts,
the random value-ordering heuristic comfortably beats the tailored strategy without restarts. In
other words, although having a good variable-ordering heuristic is beneficial, introducing
randomness into the search is better, if it is done alongside a mechanism to avoid heavy commitment
to any particular random choice. \Cref{figure:scatter-random} looks at this in more detail.  The
performance on unsatisfiable instances is the same. However, although better in aggregate, we see
many satisfiable instances performing substantially worse. ?? Quantify this a bit.

@@

\begin{figure}[p]
    \centering
    \includegraphics*{gen-graph-scatter-random.pdf}
    \caption{Instance by instance comparison of the basic algorithm with the tailored heuristic,
    versus random + restarts.}
    \label{figure:scatter-random}
\end{figure}

@@

    \caption{Instance by instance comparison of the basic algorithm with the tailored heuristic,
    versus biased + restarts.}
    \label{figure:scatter}
\end{figure}

\Cref{figure:restarts} shows that the biased heuristic together with restarts is better
still---that is, if we are introducing restarts, then it is better to add a small amount of
randomness to a tailored deterministic heuristic than it is to throw away the heuristic altogether.
In the more detailed view in \cref{figure:scatter},
nearly all of the unsatisfiable instances are on or slightly above the $x-y$ diagonal, showing that
their performance is either unchanged or reduced slightly. On the other hand, there are ?? large
numbers of satisfiable instances well below the diagonal line, indicating large speedups. Better
yet, there are only ?? satisfiable instances that are more than a factor of ?? times worse.

@@

?? maybe some engineering aspects: look at performance with restarts on tailored, and at restarts
without nogood recording

>>>>

\newcommand{\siplabel}[1]{\label{line:sip:#1}}
\newcommand{\siplineref}[1]{line~\ref{line:sip:#1}}
\begin{algorithm}[p]\DontPrintSemicolon\small\SetInd{0.1em}{1em}
    \begin{tikzpicture}[remember picture,overlay]
        \coordinate (restart1sc) at ($(pic cs:restart1s) + (0, 0.15)$);
        \coordinate (restart1ec) at ($(pic cs:restart1e) + (0, 0.00)$);
        \node [fill=magma8, rounded corners, fit=(restart1sc) (restart1ec)] { };
        \coordinate (restart2sc) at ($(pic cs:restart2s) + (0, 0.15)$);
        \coordinate (restart2ec) at ($(pic cs:restart2e) + (0, 0.00)$);
        \node [fill=magma8, rounded corners, fit=(restart2sc) (restart2ec)] { };
        \coordinate (restart3sc) at ($(pic cs:restart3s) + (0, 0.15)$);
        \coordinate (restart3ec) at ($(pic cs:restart3e) + (0, 0.00)$);
        \node [fill=magma8, rounded corners, fit=(restart3sc) (restart3ec)] { };
        \coordinate (restart4sc) at ($(pic cs:restart4s) + (0, 0.15)$);
        \coordinate (restart4ec) at ($(pic cs:restart4e) + (0, 0.00)$);
        \node [fill=magma8, rounded corners, fit=(restart4sc) (restart4ec)] { };
        \coordinate (restart5sc) at ($(pic cs:restart5s) + (0, 0.15)$);
        \coordinate (restart5ec) at ($(pic cs:restart5e) - (0, 0.10)$);
        \coordinate (restart5rc) at ($(pic cs:restart5r) + (0, 0.00)$);
        \node [fill=magma8, rounded corners, fit=(restart5sc) (restart5ec) (restart5rc)] { };
        \coordinate (restart6sc) at ($(pic cs:restart6s) + (0, 0.15)$);
        \coordinate (restart6ec) at ($(pic cs:restart6e) + (0, 0.00)$);
        \node [fill=magma8, rounded corners, fit=(restart6sc) (restart6ec)] { };
        \coordinate (restart7sc) at ($(pic cs:restart7s) + (0, 0.15)$);
        \coordinate (restart7ec) at ($(pic cs:restart7e) + (0, 0.00)$);
        \node [fill=magma8, rounded corners, fit=(restart7sc) (restart7ec)] { };
        \coordinate (restart8sc) at ($(pic cs:restart8s) + (0.05, 0.15)$);
        \coordinate (restart8ec) at ($(pic cs:restart8e) + (0, 0.00)$);
        \node [fill=magma8, rounded corners, fit=(restart8sc) (restart8ec)] { };
        \coordinate (restart9sc) at ($(pic cs:restart9s) + (0, 0.15)$);
        \coordinate (restart9ec) at ($(pic cs:restart9e) + (-0.15, 0.00)$);
        \node [fill=magma8, rounded corners, fit=(restart9sc) (restart9ec)] { };
    \end{tikzpicture}

@@

        \nl \tikzmark{restart9s}\lIf{$S \ne \KwSty{restart}$}{\tikzmark{restart9e}$\KwSty{return}~S$}

@@

@@

For value ordering, which occurs on \siplineref{valueordering} of \cref{algorithm:sip}, we will
compare five strategies:

@@

\paragraph{Degree} We select vertices from highest degree to lowest degree. This is done by
permuting the target graph at the top of search. This heuristic is the current state of the art.

\paragraph{Anti} We select vertices from lowest degree to highest degree. This heuristic is used as
a sanity check: we expect degree to beat random, and random to beat anti.

\paragraph{Position-biased} We permute the graph at the top of search, as in the degree heuristic.
Then, as with the random heuristic, we shuffle the vertices before we enter the $\KwSty{foreach}$
loop. Instead of picking a vertex uniformly at random, we select select the first (that is, highest
degree) vertex with probability 0.5, the second vertex with probability 0.25, and the $n$th with
probability $2^{-n}$.  This heuristic heavily prefers vertices of high degree, similar to the
degree heuristic, but also introduces an element of randomness, and so can be used with restarts.

\paragraph{Degree-biased} Refining position-biased, we shuffle the vertices, using a weight of
$2^{\left|\neighbourhood(v)\right|}$ when deciding whether to branch on vertex $v$ next. (We cheat
slightly to avoid range issues with random number generation: we give the vertex with highest degree
in the entire graph weight $2^{50}$, and then give lower degree vertices smaller powers, giving
every vertex weight at least 1.  This is a bit of hack but programming is hard.) This heuristic
continues to prefer vertices of high degree, but will give the same chance of being selected next to
two vertices of equal degree.

@@

In \cref{figure:value-ordering-heuristics} we show the cumulative number of instances solved over
time using the four different value-ordering heuristics. As expected, the degree heuristic is much
better (?? give aggregate speedups for both all and just satisfiable instances) than random
ordering, which is in turn ?? times better than the anti-heuristic.

@@

    \caption{An instance by instance comparison of the degree and degree-biased heuristics.}
    \label{figure:scatter-heuristics}
\end{figure}

?? Also talk about position-biased

The performance of the degree-biased and degree heuristics is very similar: the degree-biased heuristic has
slightly worse performance below a 100s timeout, and effectively equal performance above it. This
shows that we can introduce an element of randomness into the degree value-ordering heuristic
without adversely affecting its performance \emph{in aggregate}. \Cref{figure:scatter-heuristics}
gives a more detailed look at these results. In this plot, satisfiable and unsatisfiable instances
are shown using dots and crosses respectively. On the $x$-axis we measure the degree heuristic,
whilst on the $y$-axis we measure random + restarts. Points on the outer axes represent instances
which timed out with one heuristic but not the other. This plot shows that on a case by case basis,
randomness can make a large difference to individual satisfiable instances.

@@

In \cref{figure:restarts} we show the effects of adding restarts to the algorithm. With restarts,
the random value-ordering heuristic comfortably beats the degree strategy without restarts. In
other words, although having a good variable-ordering heuristic is beneficial, introducing
randomness into the search is better, if it is done alongside a mechanism to avoid heavy commitment
to any particular random choice. \Cref{figure:scatter-random} looks at this in more detail.  The
performance on unsatisfiable instances is the same. However, although better in aggregate, we see
many satisfiable instances performing substantially worse. ?? Quantify this a bit.

@@

\begin{figure}[p]
    \centering
    \includegraphics*{gen-graph-scatter-random.pdf}
    \caption{Instance by instance comparison of the basic algorithm with the degree heuristic,
    versus random + restarts.}
    \label{figure:scatter-random}
\end{figure}

@@

    \caption{Instance by instance comparison of the basic algorithm with the degree heuristic,
    versus degree-biased + restarts.}
    \label{figure:scatter}
\end{figure}

\Cref{figure:restarts} shows that either biased heuristic together with restarts is better
still---that is, if we are introducing restarts, then it is better to add a small amount of
randomness to a tailored deterministic heuristic than it is to throw away the heuristic altogether.
The degree-biased heuristics is stronger than position-biased, showing that exploiting ties and
relative sizes in the heuristic ordering is also worthwhile.

In the more detailed view in \cref{figure:scatter}, comparing the basic algorithm with the degree
heuristic to the degree-biased algorithm with restarts, nearly all of the unsatisfiable instances
are on the $x-y$ diagonal, showing that their performance is unchanged. On the other hand, there are
?? large numbers of satisfiable instances well below the diagonal line, indicating large speedups.
Better yet, there are only ?? satisfiable instances that are more than a factor of ?? times worse.

@@

?? maybe some engineering aspects: look at performance with restarts on degree, and at restarts
without nogood recording
